Finite-state morphological transducers for three Kypchak languages

### 2014-03-14 kipchak-lrec-2013/kipchak-paper.tex
some minor fixes
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@50741 72bbbca6-d526-0410-a7d9-f06f51895060

\scalebox{0.80}{
\begin{tabular}{lll}
 \textbf{Kazakh} & \textbf{Tatar} & \textbf{Kumyk} \\
\hline
 Құдай Өзінің жаратқандарының  &  Аллаһ Үзе яраткан нәрсәләргә карап, & Аллагь Оьзю яратгъан затлагъа\\
 бәріне қарап, өте жақсы екенін көрді. &  аларның бик яхшы икәнен күрде. & къарап, олар бек яхшы экенин гёрген. \\
\hline
 Құдай\tags{<n><nom>} & Аллаһ\tags{<n><nom>} & Аллагь\tags{<n><nom>}  \\
 Өз\tags{<prn><ref><px3sp><gen>} & Үз\tags{<prn><ref><px3sp><nom>} & Оьз\tags{<prn><ref><px3sp><nom>} \\
 жарат\tags{<v><tv><ger\_past><pl><px3sp><gen>} & ярат\tags{<v><tv><gpr\_past>} & ярат\tags{<v><tv><gpr\_past>} \\
 бәрі\tags{<prn><qnt><px3sp><dat>} & нәрсә\tags{<prn><itg><pl><dat>} & зат\tags{<n><pl><dat>} \\
 қара\tags{<v><tv><gna\_perf>} & кара\tags{<v><tv><gna\_perf>} & къара\tags{<v><tv><gna\_perf>} \\ 
,\tags{<cm>} & & ,\tags{<cm>} \\
 ---      & алар\tags{<prn><pers><p3><pl><gen>} & олар\tags{<prn><pers><p3><pl><nom>} \\
 өт\tags{<adv>} & бик\tags{<adv>} & бек\tags{<adv>} \\
 жақсы\tags{<adj>} & яхшы\tags{<adj>} & яхшы\tags{<adj>} \\
 е\tags{<cop><ger\_past><px3sp><acc>} & и\tags{<cop><ger\_past><px3sp><acc>} & э\tags{<cop><ger\_past><px3sp><acc>} \\
 көр\tags{<v><tv><ifi><p3><sg>} & күр\tags{<v><tv><past><p3><sg>} & гёр\tags{<v><tv><past><p3><sg>} \\  
 .\tags{<sent>} & .\tags{<sent>} & .\tags{<sent>} \\
\hline
\end{tabular}
}
\label{table:exoutput}
\caption{An example of the output of each of the morphological transducers for the same 
  sentence (``And God saw every thing that he had made, and, behold, it was 
  very good.'', Genesis 1:31). The sentence may be very roughly glossed as ``God own created things to looking,
  their good being saw''. The output has been 
  abbreviated to only show the appropriate tag in context, 
  the actual output would be ambiguous.}

@@

Most of the stems in the lexicon can be used, with no extra suffixes in any of these functions, for 
example the adjective жақсы in Kazakh can be used as attributively, жақсы китап `good book', adverbially
жақсы ойнадыңыз `(you) played well', and substantively (as a noun) `жақсы келетін еді' `(the) good (ones) would come'. This
is a productive process and relevant to the morphotactics as an adjective used substantively may take the 
full range of case and possessive suffixes, and adjectives used adverbially will take a different set of 
clitics to adjectives used substantively and attributively.

@@

Regarding numerals, work has been done for Finnish in the finite-state framework \cite{karttunen06}, however
this relies on converting all numerals to their fully spelt out form, which would involve complex operations 
on the transducer.

% FIXME: cyrillic in \texttt
Our solution is to add phonological information at the end of morphemes that need it in the form of special ``abstract letters'' that trigger phonological processes at the morphophonological stage and are deleted. For example, the string \texttt{5<num><subst><abl>} would have
the morphotactic representation 5-\{э\}\{с\}>\{D\}\{A\}н, where \{э\} and \{с\} stand for phonological triggers, and \{D\}\{A\}н is the morphotactic representation of the ablative. The symbol \{э\} signals front harmony, and \{с\} signals final unvoiced consonant. So, the rules that operate on \{D\} and \{A\} will be able to do the forward voicing assimilation \{D\} surfacing as т /t/ and the front harmony \{A\} surfacing as е /e/.

>>>>

\scalebox{0.80}{
\begin{tabular}{lll}
 \textbf{Kazakh} & \textbf{Tatar} & \textbf{Kumyk} \\
\hline
 Құдай Өзінің жаратқандарының  &  Аллаһ Үзе яраткан нәрсәләргә карап, & Аллагь Оьзю яратгъан затлагъа\\
 бәріне қарап, өте жақсы екенін көрді. &  аларның бик яхшы икәнен күрде. & къарап, олар бек яхшы экенин гёрген. \\
\hline
 Құдай\tags{<n><nom>} & Аллаһ\tags{<n><nom>} & Аллагь\tags{<n><nom>}  \\
 Өз\tags{<prn><ref><px3sp><gen>} & Үз\tags{<prn><ref><px3sp><nom>} & Оьз\tags{<prn><ref><px3sp><nom>} \\
 жарат\tags{<v><tv><ger\_past><pl><px3sp><gen>} & ярат\tags{<v><tv><gpr\_past>} & ярат\tags{<v><tv><gpr\_past>} \\
 бәрі\tags{<prn><qnt><px3sp><dat>} & нәрсә\tags{<prn><itg><pl><dat>} & зат\tags{<n><pl><dat>} \\
 қара\tags{<v><tv><gna\_perf>} & кара\tags{<v><tv><gna\_perf>} & къара\tags{<v><tv><gna\_perf>} \\ 
,\tags{<cm>} & & ,\tags{<cm>} \\
 ---      & алар\tags{<prn><pers><p3><pl><gen>} & олар\tags{<prn><pers><p3><pl><nom>} \\
 өте\tags{<adv>} & бик\tags{<adv>} & бек\tags{<adv>} \\
 жақсы\tags{<adj>} & яхшы\tags{<adj>} & яхшы\tags{<adj>} \\
 е\tags{<cop><ger\_past><px3sp><acc>} & и\tags{<cop><ger\_past><px3sp><acc>} & э\tags{<cop><ger\_past><px3sp><acc>} \\
 көр\tags{<v><tv><ifi><p3><sg>} & күр\tags{<v><tv><past><p3><sg>} & гёр\tags{<v><tv><past><p3><sg>} \\  
 .\tags{<sent>} & .\tags{<sent>} & .\tags{<sent>} \\
\hline
\end{tabular}
}
\label{table:exoutput}
\caption{An example of the output of each of the morphological transducers for the same 
  sentence (``And God saw every thing that he had made, and, behold, it was 
  very good.'', Genesis 1:31). The sentence may be very roughly glossed as ``God own-his created things/everything-at looking,
  their very good being saw''. The output has been 
  abbreviated to only show the appropriate tag in context, 
  the actual output would be ambiguous.}

@@

Most of the stems in the lexicon can be used, with no extra suffixes in any of these functions, for 
example the adjective жақсы in Kazakh can be used as attributively, жақсы кітап `good book', adverbially
жақсы ойнадыңыз `(you) played well', and substantively (as a noun) `жақсы келетін еді' `(the) good (ones) would come'. This
is a productive process and relevant to the morphotactics as an adjective used substantively may take the 
full range of case and possessive suffixes, and adjectives used adverbially will take a different set of 
clitics to adjectives used substantively and attributively.

@@

Regarding numerals, work has been done for Finnish in the finite-state framework \citep{karttunen06}; however, this relies on converting all numerals to their fully spelt out form, which would involve complex operations 
on the transducer.

% FIXME: cyrillic in \texttt
Our solution is to add phonological information at the end of morphemes that need it in the form of special ``abstract letters'' that trigger phonological processes at the morphophonological stage and are deleted. For example, the string \texttt{5<num><subst><abl>}\footnote{The resulting forms, 5-тен\super{\texttt{kaz}}, 5-тән\super{\texttt{tat}}, and 5-ден\super{\texttt{kum}} would be pronounced бестен\super{\texttt{kaz}}, биштән\super{\texttt{tat}}, and бешден\super{\texttt{kum}}.} would have
the morphotactic representation 5-\{э\}\{с\}>\{D\}\{A\}н, where \{э\} and \{с\} stand for phonological triggers, and \{D\}\{A\}н is the representation of the ablative morpheme at the morphotactic level. The symbol \{э\} signals that the following vowel needs to harmonise to a front unrounded vowel, and \{с\} signals that there is a final voiceless consonant.  So, the rules that operate on \{D\} and \{A\} will be able to do the forward voicing assimilation \{D\} surfacing as т /t/ and the front harmony \{A\} surfacing as е /e/.
